Automatically Splitting a Staged Lambda Calculus

### 2015-07-10 paper/splitting.tex
splitting reorganization

@@

Top level programs in \lang\ are given by a term $\coltwo e A$.
Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
We write this as $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.

Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
This output is evaluated in two passes by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

The correctness of the splitting translation requires that this simple 
evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

\TODO ADD EQUIVALENCE FIGURE

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on its first stage components.
Analogously, this must hold of the output of splitting, 
which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.

Thus $i$ and $q$ embody the complementary portions of $v$ that are relevant to each stage.
For this reason we call $i$ the \emph{immediate value} and $q$ the residual.

@@

For any $\colmix e A$, 
splitting $e$ (via $\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$.

The responsibilities of this first stage component, $c$, can be seen as a combination of the
those of world \bbtwo\ term splitting and partial value splitting.
Since $e$ is a term, the first stage computations within it may yet still
affect its second stage computations,
and so $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$).
But since $e$ is a world \bbonem\ construct, it has an eventual result at the 
first stage as well as the second, and so it must produce an immediate value $i$.
The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
and so we call it a {\em combined term}.

>>>>

\subsection{Outputs of Splitting}

Splitting a top level program $\coltwo e A$, via $\splittwosym$, yields $\pipeS p l r$.
Like $e$, this output is evaluated in two passes.
The first pass reduces $p$ to the value $b$ and plugs this result in for $l$ to produce $[b/l]r$;
the second pass evaluates $[b/l]r$.
The relationship between the two stages in this case is thus like a pipeline,
which is why we write them with a `\mytt{|}' in between.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

\TODO prior work statement

Splitting a partial value $v$, via $\vsplito$, yields $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on its first stage components.
Analogously, this must also hold for the output of splitting, 
which is why $q$---unlike the resumer of world \bbtwo\ term splitting---is not open on a variable.
Thus, $i$ and $q$ are operationally independent, but they represent the complementary portions of $v$ that are relevant to each stage.
We call $i$ the \emph{immediate value} and $q$ the residual.

For any $\colmix e A$, splitting $e$ via $\splitonesym$ yields the pair of monostage terms $\pipeM c l r$. 
This output form is essentially a hybrid of the previous two.
Because $e$ is a term, $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$)
And since $e$ types at world \bbonem, it has an eventual result at the 
first stage as well as the second, and so it must produce an immediate value $i$.
The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
and so we call it a \emph{combined term}.

@@

The correctness of the splitting translation requires that the simple 
evaluate-and-plug semantics on the output produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

@@
